---
title: "Introduction"
author: "Fill In Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes: |
   \setbeamertemplate{footline}{\begin{beamercolorbox}{section in head/foot}
   \includegraphics[height=.5cm]{../../Resources/Images/egap-logo.png} \hfill
   \insertframenumber/\inserttotalframenumber \end{beamercolorbox}}
output: 
  beamer_presentation:
    keep_tex: true
    toc: true
  revealjs::revealjs_presentation:
    theme: default
    highlight: pygments
    center: false
    transition: fade
    smart: false
    self_contained: false
    reveal_plugins: ["notes", "search", "chalkboard"]
    reveal_options:
      slideNumber: true
      previewLinks: true
      chalkboard:
        theme: whiteboard
        toggleNotesButton: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

htmltools::img(src = knitr::image_uri(file.path(R.home("doc"), "html", "logo.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')

```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

