---
title: "Ten tests, no effect: an introduction to multiple testing"
author: "Jake Bowers --- EGAP Learning Days, Santiago de Chile"
date: today
lang: en
bibliography: ../assets/learningdays-book.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup}
#| include: false
library(knitr)
library(coin)
library(estimatr)
opts_chunk$set(echo = TRUE, message = FALSE, comment = NA)
options(digits = 3, scipen = 3)
```

## The situation

Imagine we evaluate a social program in 100 households: half, assigned at
random, receive the program. Because the program could affect many dimensions
of life, we measure ten outcome variables: income, employment, school
attendance, health, trust in neighbors, civic participation, savings,
perceived security, consumption, and life satisfaction. At the end of the
study we run one hypothesis test per outcome and report which ones produced a
p-value below 0.05.

Each individual test promises to err rarely: if the program has no effect on
that outcome, the probability of declaring a "significant effect" by pure
chance is 5%. This document asks a different question: if the program has no
effect on **any** of the ten outcomes, what is the probability that **at
least one** of the ten tests declares an effect? Intuition says "5%, same as
before." The correct answer is about 40%.

To see this, we will build a world where we know for certain that the program
does nothing --- because we generate the data ourselves --- and then look at
what the tests produce in that world.

## A world where the program does nothing

Each household has, for each variable, two **potential outcomes**: the
outcome it would show if it received the program, $y(1)$, and the outcome it
would show if it did not, $y(0)$. "The program does nothing" has an exact
translation into this language: $y(1) = y(0)$ for every household and every
outcome. We build that world directly:

```{r data}
set.seed(20260751) ## master seed: fixes Y0 and the first experiment's assignment

N <- 100 ## households
K <- 10 ## outcome variables

outcome_names <- c(
  "income", "employment", "attendance", "health", "trust",
  "participation", "savings", "security", "consumption", "satisfaction"
)

## Potential outcomes: Y0 under control, Y1 under treatment.
## "The program does nothing" is literally Y1 = Y0,
## for every household and every outcome.
Y0 <- matrix(rnorm(N * K), nrow = N, dimnames = list(NULL, outcome_names))
Y1 <- Y0
```

The null hypothesis --- the program changes nothing for anyone --- is true by
construction, for all ten outcomes at once. Now we assign the treatment at
random (50 treated households, 50 control) and observe, for each household,
the potential outcome that corresponds to its assignment: $Y_{obs} = Z \cdot
y(1) + (1 - Z) \cdot y(0)$.

```{r assignment}
Z <- sample(rep(c(0, 1), N / 2)) ## completely randomized assignment

## We observe Y1 in the treated and Y0 in the controls.
## Because Y1 = Y0, the assignment changes nothing about what we see.
Yobs <- Z * Y1 + (1 - Z) * Y0
dat <- data.frame(Z = factor(Z), Yobs)
```

## One experiment, ten tests

We do what the evaluation team would do: one test per outcome. We use two
different procedures to show that the problem does not come from any
particular test. `oneway_test()` from the `coin` package compares means using
the permutation distribution: it re-shuffles the assignment 5000 times and
asks how extreme the observed statistic is among those re-shuffles.
`lm_robust()` from the `estimatr` package estimates the difference in means
with HC2 standard errors and a normal approximation. Both are justified by
the random assignment --- the first uses it directly, the second
approximates it.

```{r oneexperiment}
p_oneway <- sapply(outcome_names, function(v) {
  pvalue(oneway_test(reformulate("Z", v),
    data = dat,
    distribution = approximate(nresample = 5000)
  ))
})

p_lm <- sapply(outcome_names, function(v) {
  est <- tidy(lm_robust(reformulate("Z", v), data = dat))
  est$p.value[est$term == "Z1"]
})

table1 <- data.frame(
  outcome = outcome_names,
  p_oneway = round(p_oneway, 3),
  p_lm_robust = round(p_lm, 3),
  significant = ifelse(p_lm <= 0.05, "*", "")
)
kable(table1, row.names = FALSE)
```

Look at the `health` row: p = `{r} round(p_lm[["health"]], 3)` with both
tests.^[We chose the master seed from among several so that this first
experiment would fall inside the ~40% of experiments that produce at least
one false positive. The simulation below shows how often that happens; it is
not a rare case.] If this were a real study, the report would say "the
program improved health (p < 0.05)" --- and perhaps the executive summary
would mention only that result. But we know the program did nothing: we built
`health` with $y(1) = y(0)$ for every household. The two tests agree almost
exactly, because the problem is not in the test. It is in how many questions
we asked the data.

## The family-wise error rate (FWER)

First in words. When we run a single test at significance level
$\alpha = 0.05$, the 5% is the probability of a false positive **in that
test**. When we run ten tests, the set of ten is a **family**, and the
question that matters to the reader of the report is: what is the probability
that the whole family produces **at least one** false positive? That
probability is called the **family-wise error rate** (FWER).

Now with a little algebra. Suppose the $K$ tests were independent and all the
null hypotheses were true. Each test "gets it right" (does not reject) with
probability $1 - \alpha$: with $\alpha = 0.05$, it gets it right 95% of the
time. With what probability do the first two get it right at the same time?
Because they are independent, we multiply:
$0.95 \times 0.95 \approx 0.90$. Each additional test multiplies by 0.95
again, so after $K$ tests the probability that **all** of them get it right
is $(1 - \alpha)^K$ --- a number that shrinks with every test we add. And "at
least one false positive" is exactly the opposite of "all get it right", so
its probability is one minus that:

$$
\text{FWER} = 1 - (1 - \alpha)^K.
$$

With $\alpha = 0.05$ and $K = 10$, that gives $1 - 0.95^{10}$ =
`{r} round(1 - 0.95^10, 3)`. Back to words: a team that evaluates ten
outcomes of a program with no effect has about a 40% chance of "discovering"
something. With twenty outcomes, the probability reaches
`{r} round(1 - 0.95^20, 2)`. The curve grows fast:

```{r fwercurve}
#| echo: false
#| fig-width: 6
#| fig-height: 4
K_seq <- 1:20
par(mar = c(4, 4, 1, 1), mgp = c(2.2, 0.7, 0))
plot(K_seq, 1 - (1 - 0.05)^K_seq,
  type = "b", pch = 19, ylim = c(0, 0.7),
  xlab = "Number of tests (K)",
  ylab = "Prob. of at least one false positive"
)
abline(h = 0.05, lty = 2)
text(18.5, 0.08, expression(alpha == 0.05))
```

In a real experiment the outcomes are usually correlated with each other, and
then the formula $1 - (1-\alpha)^K$ is only an approximation. That is why we
do not stop at the algebra: we simulate.

## The simulation: we repeat the experiment 1000 times

In our constructed world, the only source of randomness is the treatment
assignment --- exactly the source of randomness that justifies the two tests.
So we keep the potential outcomes `Y0` and `Y1` fixed and re-draw the
assignment $B = 1000$ times, applying the same observation equation each
time. In each repetition we run the ten tests with each procedure and store
the p-values.

```{r simulation}
#| cache: true
run_experiment <- function() {
  ## Re-draw the assignment and observe again; Y0 and Y1 stay fixed.
  Z_new <- sample(rep(c(0, 1), N / 2))
  d <- data.frame(Z = factor(Z_new), Z_new * Y1 + (1 - Z_new) * Y0)
  p_ow <- sapply(outcome_names, function(v) {
    pvalue(oneway_test(reformulate("Z", v),
      data = d,
      distribution = approximate(nresample = 5000)
    ))
  })
  p_lm <- sapply(outcome_names, function(v) {
    est <- tidy(lm_robust(reformulate("Z", v), data = d))
    est$p.value[est$term == "Z1"]
  })
  rbind(oneway = p_ow, lm_robust = p_lm)
}

set.seed(20260731) ## simulation seed
B <- 1000
sims <- replicate(B, run_experiment()) ## dimensions: 2 procedures x K x B
```

Before looking at the FWER, let us verify that each individual test keeps its
promise. For each outcome, we compute the proportion of the 1000 repetitions
in which that test rejected with $p \leq 0.05$:

```{r pertest}
reject_ow <- rowMeans(sims["oneway", , ] <= 0.05)
reject_lm <- rowMeans(sims["lm_robust", , ] <= 0.05)
kable(
  data.frame(
    outcome = outcome_names,
    oneway_test = round(reject_ow, 3),
    lm_robust = round(reject_lm, 3)
  ),
  row.names = FALSE
)
```

Each number estimates a proportion with $B = 1000$ repetitions, so it carries
a simulation error (Monte Carlo standard error) of
$\sqrt{0.05 \times 0.95 / 1000} \approx `{r} round(sqrt(.05 * .95 / 1000), 3)`$.
All the rates are where they should be: near 5%. Test by test, nothing is
broken.

Now the family question: in what proportion of the 1000 repetitions did **at
least one** of the ten tests reject?

```{r fwer}
fwer_ow <- mean(apply(sims["oneway", , ] <= 0.05, 2, any))
fwer_lm <- mean(apply(sims["lm_robust", , ] <= 0.05, 2, any))
c(oneway_test = fwer_ow, lm_robust = fwer_lm)
```

The estimated FWER is `{r} round(fwer_ow, 2)` with `oneway_test()` and
`{r} round(fwer_lm, 2)` with `lm_robust()` (Monte Carlo standard error
$\approx `{r} round(sqrt(.4 * .6 / 1000), 3)`$), consistent with the
theoretical value of `{r} round(1 - 0.95^10, 3)` for independent tests. This
is the contrast that matters: **every row of the previous table says 5%, and
yet the family errs 40% of the time.** In four of every ten evaluations of a
program that does nothing, the report can say "we found a significant effect"
without anyone having cheated.

## The correction: one line of code

Multiple-comparison procedures adjust the p-values so that the 5% promise
applies to the whole family and not to each test separately [@bretz2011]. The
simplest is Bonferroni: compare each p-value with $\alpha / K$ instead of
$\alpha$. The Holm procedure does the same thing step by step and rejects at
least as much as Bonferroni, so we use Holm. In R it is one line:
`p.adjust(p, method = "holm")`.

We repeat the FWER calculation, now adjusting the ten p-values within each
repetition before asking whether any falls below 0.05:

```{r holm}
fwer_holm_ow <- mean(apply(sims["oneway", , ], 2, function(p) {
  any(p.adjust(p, method = "holm") <= 0.05)
}))
fwer_holm_lm <- mean(apply(sims["lm_robust", , ], 2, function(p) {
  any(p.adjust(p, method = "holm") <= 0.05)
}))
c(oneway_test = fwer_holm_ow, lm_robust = fwer_holm_lm)
```

With the Holm adjustment, the estimated FWER drops to
`{r} round(fwer_holm_ow, 3)` and `{r} round(fwer_holm_lm, 3)` --- below the
promised 5%. The correction is not free: demanding smaller p-values means
less power to detect real effects when they exist. That cost is a reason to
declare few primary hypotheses, or to group outcomes into indices, rather
than test everything that was measured. When the family is very large,
controlling a different rate --- the expected proportion of false discoveries
among the rejections (FDR) --- can be a more reasonable target than the FWER
[@benjamini1995].

## What does this have to do with transparency?

The 40% calculation assumes something that almost never holds: that the
reader sees all ten tests. If the report mentions only `health`, the reader
cannot know that it was one test among ten, and the p-value of 0.01 looks
like strong evidence. Nor does the problem require bad faith: it is enough
for the analysis decisions --- which outcomes to test, how to define them,
which subgroups to examine --- to be made after seeing the data
[@gelman2013garden].

The pre-analysis plan answers exactly this. Declaring in advance which
hypotheses are primary, how each outcome is defined, and which
multiple-testing correction will be applied makes the size of the family
visible. The FWER guarantee is a promise about the complete family; it only
means something to a reader who can see the complete family. That is what
registering the design delivers.

## References
