---
title: "Estimation, ATE, SE"
author: "Natalia Garbiras-Díaz"
date: "April 10, 2019"
output:
  slidy_presentation:
      fig_height: 4
  beamer_presentation:
    keep_tex: yes
    toc: yes
  revealjs::revealjs_presentation:
    center: no
    highlight: pygments
    reveal_options:
      chalkboard:
        theme: whiteboard
        toggleNotesButton: no
      previewLinks: yes
      slideNumber: yes
    reveal_plugins:
    - notes
    - search
    - chalkboard
    self_contained: no
    smart: no
    theme: default
    transition: fade
---



A simulation in R: sample mean as an unbiased estimator of the population mean
======================================================================

First, we will need to "create" a population (a study group)

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(strip.white=TRUE,
               width.cutoff=132,
               size='\\scriptsize',
               out.width='.9\\textwidth',
               message=FALSE,
	       warning=FALSE,
	       echo=TRUE,
               comment=NA,
               tidy='styler',
               prompt=FALSE,
	       results='markup')
```


```{r}
population <- c(4,5,7,12,7,8,9,-3,5,8,9,3,2,3,4,6,10,4,6,7,8,9,2)

N <- length(population) # number of observations in the population
N

pop_mean <- mean(population) # population mean
pop_mean

```

----

We will draw several random samples of 8 observations ($m$) each *without* replacement

```{r}

set.seed(12345)
s1 <- sample(population, size=8, replace = FALSE)

s2 <- sample(population, size=8, replace = FALSE)

s3 <- sample(population, size=8, replace = FALSE)

s4 <- sample(population, size=8, replace = FALSE)

samples <- rbind(s1, s2, s3, s4)

samples
```

----

Remember the population mean: `r pop_mean`

And the means of the four samples

```{r}
apply(samples, MARGIN=1, FUN=mean) # apply function to rows
```

By chance each given sample mean may be a little higher or lower than the population mean.

We can use R to show that the sample mean is an unbiased estimator of the population mean.

----

For this, we will write a *simulation*. We will repeat the sampling process $10,000$ times.

```{r}

sample_mean <- NA

for (i in 1:10000){

  sample <- sample(population, size=8, replace = FALSE)
  sample_mean[i] <- mean(sample)

}

```

----

```{r}
par(mfrow=c(1,1))
plot(density(sample_mean), col="blue", lwd=3,
     main="Distribution of sample means")
abline(v=pop_mean, col="red", lwd=2)

average_sampling_distribution<- mean(sample_mean)
average_sampling_distribution
round(pop_mean, 4)

```

Let's now look at the distribution of the sample mean as $m$ gets closer to N.
======================================================================

So far, $m=8$. We now need a new simulation that adds a new step: we need to vary the size of *m*. (Remember our population size, *N*, is `r N`)

----

```{r, eval=FALSE}

rep <- 10000

# The first loop varies m
for (m in 9:20){

  sample_mean <- NA #creating an object to store the results of the second loop

  # The second loop goes through the 10,000 simulations
  for (i in 1:rep){

    #we first get a random sample of size m from the population
    sample <- sample(population, size=m, replace = FALSE)
    #and then calculate and store the sample mean
    sample_mean[i] <- mean(sample)
  }

  #finally, we plot the distribution of the 10,000 sample means for the relevant m
  lines(density(sample_mean), lwd=3,
        #note that this next line of code varies the color of the line according to m
        #so that we can distinguish the different distributions
        col=paste0("grey",140-(7*m)))
}

```

---

```{r, echo=FALSE}

plot(density(sample_mean), col="blue", ylim=c(0,1.6),
     main="Distribution of sample means", lwd=3)
abline(v=pop_mean, col="red", lwd=3)

rep <- 10000

for (m in 9:20){
  sample_mean_sim <- NA

  for (i in 1:rep){
    sample <- sample(population, size=m, replace = FALSE)
    sample_mean_sim[i] <- mean(sample)
  }

  lines(density(sample_mean_sim), lwd=3,
        col=paste0("grey",140-(7*m)))
}

```

The variance of the sample mean
=======================================================================

The standard deviation of the sampling distribution gives us a measure of uncertainty about the mean:

```{r}

var_sample_mean<- sum((sample_mean - mean(sample_mean))^2) / (length(sample_mean))
se_sample_mean<- sqrt(var_sample_mean)
se_sample_mean

```

Now, we can calculate this because we created our own population. This is not often the case (e.g., experiments)...

---

Remember the formula for the variance of the sample mean for the treatment group is:

$$\text{Var}(Y^T) = \frac{\sigma^2}{m}$$

We do not know $\sigma^2$, we can estimate this quantity with the variance of the assigned-to-treatment sample by:

$$\hat{\sigma}^2= (\frac{1}{m-1})\sum^{m}_{i=1}(Y_i-\bar{Y}^T)^2$$

Same with the variance of the sample mean for those units assigned to control.

2. Estimation of the ATE
======================================================================

We can write a function to estimate the ATE (or simply use the built-in function `t.test`).

```{r}

diff_means <- function(y, x){

  # Calculating difference in means
  mean1 <- mean(y[x==1], na.rm=T)
  mean0 <- mean(y[x==0], na.rm=T)
  diff <- mean1 - mean0

  # Calculating number of observations
  N <- length(na.omit(y))

  # Preparing output
  res <- c(mean1, mean0, diff, N)
  names(res) <- c("Mean 1", "Mean 0", "Difference", "N")

  return(c(res))
}
```

----

To try our function, we will use the small dataset in Gerber & Green (2012)

```{r}
gg_data <- as.data.frame(cbind(c(10,15,20,20,10,15,15),
                               c(15,15,30,15,20,15,30)))
names(gg_data) <- c("Y_i0", "Y_i1")

```

----

We will need to "create" a treatment vector...

```{r}
# let's fix m=3 (units in the treatment group)
treat <- c(1, 1, 1, 0, 0, 0, 0)
gg_data$treat <- sample(treat, 7, replace=F)
gg_data$treat

```

...and a column with the "observed" outcomes

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
```

```{r}
gg_data$observed <- ifelse(gg_data$treat==1, gg_data$Y_i1, gg_data$Y_i0)
##save(gg_data, file="gg_data.RData")

```

----

Let's see how the complete data set looks now:

```{r}
head(gg_data)
```

----

```{r}
# mean of the treatment group
mean(gg_data$observed[gg_data$treat==1])
# mean of the control group
mean(gg_data$observed[gg_data$treat==0])

# difference of means
mean(gg_data$observed[gg_data$treat==1]) - mean(gg_data$observed[gg_data$treat==0])

# with our function
diff_means(gg_data$observed, gg_data$treat)

```

---

Now, we can also estimate the same quantity using a regression:

```{r, eval=TRUE,echo=FALSE,warning=FALSE}
library(estimatr)
```

```{r}
lm_robust(observed~treat, data=gg_data)
```

But notice that we are not relying on the assumptions of OLS regression. This is just math... the way $\beta$ is estimated.

----

How can we get a distribution of the difference of means?
==============================================

We can do this with a simulations. For each simulation,

> - First: We will need to "create" a random treatment vector and generate the column with the associated observed outcomes.

> - Second: We will have to calculate the difference between the treatment and control means (by hand or using our new function).

---

```{r}

# 1.
gg_data$treat <- sample(treat, 7, replace=F)
gg_data$observed <- ifelse(gg_data$treat==1, gg_data$Y_i1, gg_data$Y_i0)

# 2.
diff_means(gg_data$observed, gg_data$treat)
# we should store this! so,
dm <- diff_means(gg_data$observed, gg_data$treat)
dm
# but we only want the third element!
dm <- diff_means(gg_data$observed, gg_data$treat)[3]
dm

```

----

Now let's put this in a loop that allows us to repeat the process $10,000$ times (and saves the dom for each)...

```{r}

dm <- NA #creating a placeholder to store all our doms...

for (i in 1:10000){

    # 1.
    gg_data$treat_sim <- sample(treat, 7, replace=F)
    gg_data$observed <- ifelse(gg_data$treat_sim==1, gg_data$Y_i1, gg_data$Y_i0)

    # 2.
    dm[i] <- diff_means(gg_data$observed, gg_data$treat_sim)[3]

    }

```

----

Finally, let's plot the distribution

```{r}
hist(dm, col="blue", main="Histogram of Difference of Means \n for GGdata")
```


3. Standard Error for the ATE
=====================================================================

**1. Standard error for the difference in means**

```{r, echo=FALSE}
library(foreign) # to import csv
library(ggplot2) # for plots using ggplot
```

---

1. Standard error for the difference in means
==========================================================

+ The difference in means is an unbiased estimator of the true ATE. However, by chance, in some realizations of our sample that estimate might be off the true ATE.

+ The SE tells us the likely size of the amount off.

A conservative formula for the $\widehat{SE}$ for the $\widehat{ATE}$
================================================

$$\widehat{SE}(\widehat{ATE})= \sqrt{\frac{\widehat{\text{Var}}(Y_i(0))}{N-m} + \frac{\widehat{\text{Var}}(Y_i(1))}{m}}$$

---

We are going to estimate the SE for the difference in means using the same data.

```{r, echo=FALSE}
load("gg_data.RData")
```

```{r}
true_ate<-mean(gg_data$Y_i1) - mean(gg_data$Y_i0)
true_ate
est_ate<- mean(gg_data$observed[gg_data$treat==1]) - mean(gg_data$observed[gg_data$treat==0])
est_ate

```

---

```{r}

# generating empty dataframe to put the results
ate <- as.data.frame(matrix(NA, 10000, 2))
names(ate) <- c("estimated_ate", "estimated_se_ate")

# sampling
for (i in 1:10000){

  # generating treatment vector for this replicate
  gg_data$treat_sim <- 0
  gg_data$treat_sim[sample(1:7, 2, replace=F)]  <- 1

  treat_mean <- mean(gg_data$Y_i1[gg_data$treat_sim==1])
  treat_var <- var(gg_data$Y_i1[gg_data$treat_sim==1])

  control_mean <- mean(gg_data$Y_i0[gg_data$treat_sim==0])
  control_var <- var(gg_data$Y_i0[gg_data$treat_sim==0])

  ate[i,1] <- treat_mean - control_mean
  ate[i,2] <- sqrt(treat_var/2 + control_var/5)
}

```

---

```{r, echo=FALSE}

m <- ggplot(ate, aes(x=estimated_ate))
m +
  geom_histogram(aes(y = 100 * (..count..)/sum(..count..)), binwidth=.5, alpha=.5) +
  # geom_histogram(aes(y = ..density..)) +
  geom_vline(xintercept = mean(ate$estimated_ate), col="red", size=1.25) +
  theme_bw() +
  xlab("Estimated ATE") +
  ylab("Percent")


```

+ How could we use this graph to get the SE of the estimated ATE?

---

```{r}
# The SE of the estimated ATE is the standard deviation of this sampling distribution:
se_sampling<-sd(ate[,1])
se_sampling

```

---

+ But in any given experiment, we don't have the sampling distribution. Instead, we can estimate the SE (using the conservative formula)

```{r}

treat_var <- var(gg_data$Y_i1[gg_data$treat==1])
control_var <- var(gg_data$Y_i0[gg_data$treat==0])
est_se_cons<- sqrt(treat_var/2 + control_var/5)
est_se_cons

```

---

```{r}

# Comparing the true standard error to the conservative formula
print(c(se_sampling, est_se_cons))

```

4. Blocked randomized experiments
=====================================================================

```{r,echo=FALSE,warning=FALSE,message=FALSE}

# We need
library(randomizr)
library(dplyr)
rm(list = ls())
set.seed(12345)

villages <- c("vill 01","vill 02","vill 03","vill 04","vill 05",
              "vill 06","vill 07","vill 08","vill 09","vill 10")

samples <- c(60,60,60,60,60,
             60,60,60,60,60)

N <- sum(samples)
ID <- 1:N

village <- rep(x = villages, times = samples)

female <- rep(c(rep(1,30),rep(0,30)),10)
days.sick.no.device <- rnbinom(n = N,mu = 10,size = 1) + 7

outbreak.effect <- 5
outbreak.villages <- sample(x = villages,size = 3)
Y0 <- ifelse(test = village %in% outbreak.villages,
             yes = days.sick.no.device + outbreak.effect,
             no = days.sick.no.device + 0)

effect.male <- -2
effect.female <- -7
Y1 <- ifelse(test = female == 1,yes = Y0 + effect.female, no = Y0 + effect.male)

data <- data.frame(
     ID = ID,
     village = village,
     female = female,
     Y0 = Y0,
     Y1 = Y1
)

complete.rand <- c(rep(1,200),
                 rep(0,N-200))

complete.rand <- sample(complete.rand)
data$complete.rand <- complete.rand
data$complete.obs <- with(data,Y1*complete.rand+Y0*(1-complete.rand))
```

---

Let's use the data from yesterday, with the example of water sanitazing devices. We had

```{r}

table(data$complete.rand,data$female)

# Block randomization using randomizr

data$block.rand <- block_ra(blocks = data$female, prob_each=c(.75,.25))

table(data$block.rand,data$female)

data$block.obs <- with(data,Y1*block.rand+Y0*(1-block.rand))

head(data)

```


How can we analyze these data?
=====================================================================

+ When analyzing data from blocked randomized experiments, we may ask different questions:
    + For instance, what is the ATE among women? Does the ATE vary by gender?
    + We may, instead, be interested in the overall ATE.

---

+ Since we conducted a complete RA at the block level, we can estimate the ATE for each one of the groups created by our blocking variables

```{r, eval=FALSE}
#Recall
effect.male <- -2
effect.female <- -7
```


```{r}
female<-filter(data, data$female==1)
dom_fem<-mean(female$block.obs[female$block.rand==1])- mean(female$block.obs[female$block.rand==0])
dom_fem

male<-filter(data, data$female==0)
dom_male<-mean(male$block.obs[male$block.rand==1])- mean(male$block.obs[male$block.rand==0])
dom_male

```

---

+ Now, we can also estimate the overall ATE by estimating block-level ATEs.
+ We then need to ask, how do we want to weight each block-level ATE in order to obtain the overal ATE?
+ One way is to weight by the block size:

```{r}

block_female<-  sum(data$female==1)/length(data$ID)
block_male<-  sum(data$female==0)/length(data$ID)

ate_overall<- block_female*dom_fem + block_male*dom_male
ate_overall
var_fem_treat <- var(data$block.obs[data$block.rand==1 & data$female==1])
var_fem_control <- var(data$block.obs[data$block.rand==0 & data$female==1])
var_male_treat <- var(data$block.obs[data$block.rand==1 & data$female==0])
var_male_control <- var(data$block.obs[data$block.rand==0 & data$female==0])

se_est_fem<-sqrt(var_fem_control/sum(data$block.rand==0 & data$female==1) + var_fem_treat/sum(data$block.rand==1 & data$female==1))
se_est_male<-sqrt(var_male_control/sum(data$block.rand==0 & data$female==0) + var_male_treat/sum(data$block.rand==1 & data$female==0))

se_est_all<- sqrt((se_est_fem^2*block_female^2) + (se_est_male^2*block_male^2))
se_est_all

```

---

We could have done this using the `difference_in_means` command from `estimatr`

```{r}
difference_in_means(block.obs~block.rand, blocks=female,data=data)
(c(ate_overall,se_est_all))

```

---

Imagine we forget that we blocked:

```{r}
lm_robust(block.obs~block.rand, data=data)
(c(ate_overall,se_est_all))

```

---

We could also get to this quantity using a regression with block dummies (Least Squares Dummy Variables) or with weights (IPW):

```{r}

table(data$block.rand, data$female)

# Block dummies (LSDV):
# the weigths used here are: p_j * (1 - p_j) * n_j)
lm_robust(block.obs~block.rand + female, data=data)

(c(ate_overall,se_est_all))

within_block <-
  data %>%
  group_by(female) %>%
  summarise(block_ATE_est = mean(block.obs[block.rand == 1]) - mean(block.obs[block.rand == 0]),
            n_j = n(),
            p_j = mean(block.rand),
            sample_weight = n_j,
            fe_weight     = p_j * (1 - p_j) * n_j) %>%
  # divide by the sum of the weights
  mutate(sample_weight = sample_weight/sum(sample_weight),
         fe_weight = fe_weight/sum(fe_weight))

head(within_block)

within_block %>%
  summarize(LDSV = weighted.mean(block_ATE_est, fe_weight),
            Blocked_DIM = weighted.mean(block_ATE_est, sample_weight))

# IPW:

data$cond_prob = ifelse(data$block.rand==1, 0.25, 0.75)
lm_robust(block.obs~block.rand, weight = 1/cond_prob, data=data)
(c(ate_overall,se_est_all))

```
