#[Getting Started on R and simple statistics"](#gettingstarted)

* Getting started on R
     + What is R, how does it function?
     + Key commands
* Descriptive statistics and plotting
     + Hands-on: With your own data...

#1. Getting started on R

##1.a. What is R, how does it function?

R is an open-source (i.e., free access) programming tool. It is object-based: This means that everything in R is an object... even commands are also objects. For example:

```{r,tidy=TRUE}
mean
```

In R, you can store objects using the `<-` operator. Make sure you give names to all objects:

```{r}
my.example <- 4 + 5
my.example
```

> "Names must be unique": everytime you give an `object` a `name`, it removes anything that already had that `name` from your environment!

```{r}
my.name <- dir()
my.name
```

Whenever you are not sure about how a command works, you can pull documentation using: `?`

```{r}
?mean
```

This tool is very helpful! Familiarize with it and feel comfortable using it... 

##1.b. Key commands: 

In this short session we go over the main commands you are likely to use in these sessions. Note: (*) corresponds to more sophisticated (but very useful) commands. 

###SETTING UP YOUR R SESSION

+ First, ask R which working directory you are currently working on. Then, set your local directory: note direction of strokes / (not \\). 

```{r}
getwd()
setwd("/Users/nataliagarbirasdiaz/Dropbox/Njala University Workshop on Experiments/Workshop 2016 (Chile)/R_handouts/Day_1_Getting_Started_R")   
```

+ Install any relevant packages (only has to be done once). 

```{r,eval=FALSE}
install.packages("Hmisc")  
```

+ *Load* any relevant packages.

```{r,message=FALSE,warning=FALSE}
library(Hmisc)
```

+ Clear R's memory     

```{r}
rm(list = ls())                                   
set.seed(20150420)        # OPT: Set a  seed to make replication possible. 
```

###BASICS ON R 

Good, you're all set to start working on R. Let's explore some of the basic commands you are going to be using these days. 

+ Creating and manipulating variables and vectors:

```{r}
a <- 5                 # "<-"  is the assignment command; it is used to define things. eg:
a

b <- 1:10              # ":"  is used to define a string of integers
b

v <- c(1,3,2,4,110,pi) # use c() to make a vector with anything in it
v

# Extract elements of a vector:

b[1]                   # Returns position 1
b[6:5]                 # returns positions 6 and 5, in that order
b[-1]                  # Returns all but the first number  

# Returns all numbers indicated as "TRUE"
b[c(TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE)]  
                                                                          
# Assign new values to particular elements of a vector
b[5] <- 0
b
```

+ Manipulating Matrices:

```{r}
matrix(1:12, 3, 4)             # Make a 3*4 matrix with numbers 1:12
matrix(1:12, 4, 3)             # Make a 4*2 matrix with numbers 1:12
matrix(1:3, 4, 3)              # Make a 4*3 matrix with numbers 1:3, cycling; filling along by column
matrix(1:3, 4, 3, byrow=TRUE)  # Make a 4*3 matrix with numbers 1:3, cycling, filling  by row
```

+ Adding row names and column names to matrices

```{r}
M<-matrix(1:12, 3, 4)
rownames(M) = c("a","b","c")
colnames(M) = c("A","B","C", "Z")
M
```

+ Simple Functions on vectors (We'll get back to these functions)

```{r}
sum(b)                     # sum
mean(b)                    # mean
max(b)                     # mean
min(b)                     # min
sd(b)                      # standard deviation
var(b)                     # variance
```

+ Simple transformations on vectors (or numbers, or matrices)

```{r}
b^2                        # Square the variable
matrix(1:3, 4, 3)^2        # Square the elements of a matrix 
b^.5                       # Square root of the variable
log(b)                     # log of variable
exp(b)                     # e to the b
```

+ Logical (asks to evaluate conditions)

```{r}
b==2                       # Is equal to
b<5                        # Less than
b>=5                       # Greater than or equal to 
b<=5 | b/4==2              # OR
b>2 & b<9                  # AND
is.na(b)                   # where is data missing
which(b<5)                 # gives indices of values meeting logical requirement
```

+ Distributions 

```{r}
rnorm(5)                      # Draws 5 obs. from normal distribution
rbinom(5, 10, .4)             # Draws from a binomial distribution
runif(5)                      # Draws from a uniform distribution
```

+ Functions on PAIRS of variables

```{r}
x  = rnorm(100)               # Create variable "x" with 100 obs. from normal distribution
y  = x+rnorm(100)

y+x                            # Add variables together (or subtract, multiply etc)
y>x                            # Logical relation between two variables?
cor(x,y)                       # Correlation between variables
t.test(y~(x>.5))               # t-test of null that variables unrelated
lm(y~x)                        # OLS regression 
M<- lm(y~x)   
summary(M)                     # Summary of OLS regression
y%*%x                          # inner product of variables

# Make a dataframe and view it
d <- data.frame(x, y)
```

+ Loops (\*): Loops repeat an operation/function over different values of *i*. See examples below:   

```{r}
x<-0                    
for(i in 1:10){         # repeat an expression for values of i in 1 to 10. 
  print(x<x+i)          # print allows you to see the results in the R console.  
   }

x=0
while(x <10){           # keep repeating an expression as long as some condition is satisfied
  print(x^2)
  x<-x+1
}

```

+ Apply and sapply (*): These two commands are very useful for efficient coding. They apply a function on an array or matrix. NOTE: You can always explore how these two commands work by typing `?apply` or `?sapply`.  

```{r}
sapply(1:4, function(j) j^2)   # faster way to do loops; this example runs function j^2 on 
                               # numbers j in 1:4
M<-matrix(rnorm(40), 4, 10)
apply(M, 1, mean)              # faster way to do lots of operations on a matrix: here take the mean 
                               # of every ROW (1)
apply(M, 2, sd)                # faster way to do lots of operations on a matrix: here take the sd  
                               # of every COLUMN (2)
```

#### Writing your own function: 

```{r}
h <- function(a, b=1) {        # Define any function by 1. assigning arguments (here a,b) and
                               # 2. giving any default values
     a^2 -b                    # (here b=1), and then giving the function
          }             
h(1)                           # illustration of function
h(1,2)                         # illustration of function

```

#### Parentheses

```{r}
# ( )   used for funtions, and for vectors: 
log(1)

# ( )   also used to keep things in blocks:
(1+2)/(2+1)
1+2/2+1

# [ ]   used to extract indices:
(10:12)[2]

# { } Used to keep blocks together, with final component returned:
x <- {
  a<-1
  b<-2
  a/b
}

x
```

#2. Descriptive statistics and ploting 

We will create two variables that are normally distributed and then will use different tools to characterize these data.

```{r}
# First, how do we do this?
y <- rnorm(10)                  # Generate y from a normal distribution 
x <- rnorm(10)                  # Generate y from a normal distribution 
```

Now, let's look at the basic descriptive statistics

```{r}

mean(y)
mean(x)                 
var(x)
sd(x)
max(x)
min(x)
summary(x)          # Basic summary statistics

cov(x,y)            # What is cov? 
mean(y[x>0])        # Estimates the mean of y when x>0

```

Now let's plot both variables. 

```{r,warning=FALSE,message=FALSE}

par(mfrow = c(1,2))             # This lets you put a set of graphs on the same canvas -- here, 2*2
hist(y)                         # A histogram
boxplot(y~(x>0))                # Box plots
dev.off()                       # We tell R that we no longer want to plot in a 1 by 2 canvas. 
```

```{r}
plot(x,y)                       # xy plots  
abline(a=-1, b = 1, col="blue") # Add a sloped line
abline(v=mean(x), col="red")    # Add a vertical line
abline(lm(y~x))                 # Add a regression line
text(0, 0.2, "some text")       # Add text
title("An exercise")            # Add a title

# A more fancy plot
x1 <- rnorm(100)
x2 <- rnorm(100) 
x3 <- rnorm(100)+x1+x2

plot(density(x1))           # We plot the distribution of x1. 

# How do we plot all three distributions?

plot(density(x1), main="Exercise")   #Notice we can add a title in the same plot command
lines(density(x2), col="slateblue")
lines(density(x3), lwd = 2)
lines(density(x1+x2), lty=2)

##Add a legend 
legend("topright", legend = c("x1","x2","x3","x1+x2"), 
       col = c("black", "slateblue", "black","black"),
       lty = c(1, 1, 1, 2), bg = "gray90")

```

##2.a. Hands-on: with your own data

The following code will help you to import data in `.csv` or `.dta` (Stata):

```{r, eval=FALSE}

#install.packages("foreign")        # This will let you import data from other formats 
                                    # Uncomment if not installed yet. 
library(foreign)

#install.packages("readstata13")    # This will let you import data from Stata 13 
                                    # Uncomment if not installed yet. 
library(readstata13)

#Stata files: 
dta<- read.dta("file_path/file_name.dta")     # Insert right directory path
dta13<- read.dta13("file_path/file_name.dta") # In case you are importing data from Stata 13

# CSV files: 
dta.csv <-dta.csv("file_path/file_name.dta") # Insert right directory path  

```

Now, fill out the following table with your data:

---------------------------------------------------------------------------------------------------
Variable                              Treatment group        Control group        Whole Sample
--------------------------------- --------------------- --------------------- ---------------------
                                   Mean  SD  Min   Max   Mean  SD  Min   Max   Mean  SD  Min   Max
**Covariates**                   

Pre-treatment outcome variable   
(if available)                   

Covariate 1                      

Covariate 2                      

**Outcome variables**            

Outcome variable 1               

Outcome variable 2               
---------------------------------------------------------------------------------------------------


Finally, use your own data to create the following plots:

+ Distribution of each one of your primary outcomes of interest
+ Plot the average for each on of your outcome variables by treatment group




